---
title: "police-involved deaths"
output:
  html_document:
    df_print: paged
---

data source: [US Police Involved Fatalities](https://data.world/awram/us-police-involved-fatalities)

```{r setup, include=FALSE}
library(tidyverse)
library(dslabs)
library(RColorBrewer)
theme_set(theme_minimal())
prim_color <- "#8fc224"
```


```{r data, include=FALSE}
# import data
# police fatality data
deaths <- readr::read_csv("./awram-us-police-involved-fatalities/data/police_fatalities.csv")
# census population data
state_pop <- readr::read_csv("./awram-us-police-involved-fatalities/data/censusstatepopulations2014.csv")
us_pop <- readr::read_csv("./awram-us-police-involved-fatalities/data/ACSDP1Y2014.DP05_data_with_overlays_2020-06-02T011259.csv")

# clean and tidy 
us_pop <- us_pop %>% 
    select(Total=DP05_0028PE,
           White=DP05_0032E,
           Black=DP05_0033E,
           Native=DP05_0034E,
           Asian=DP05_0039E,
           Hispanic=DP05_0066E
    ) %>% 
    slice(2) %>% 
    mutate(Total=as.numeric(Total),
           White=as.numeric(White),
           Black=as.numeric(Black),
           Native=as.numeric(Native),
           Asian=as.numeric(Asian),
           Hispanic=as.numeric(Hispanic)
    )
```


The data set represents data from the years **`r lubridate::year(min(as.Date(deaths$date)))`** through **`r lubridate::year(max(as.Date(deaths$date)))`**, representing **`r format(length(deaths$uid),big.mark=",")`** police-involved deaths across the United States. 

---

### reported circumstances

```{r flee, echo=FALSE}
# suspect fleeing
deaths %>% 
    count(flee) %>% 
    ggplot(aes(flee,n)) +
        geom_col(fill=prim_color,color="black",alpha=0.8,width=0.6) +
        scale_y_continuous(labels=scales::comma) +
        labs(title="fleeing?",x="",y="")
```


```{r armed, include=FALSE}
# suspect armed
deaths %>% 
    count(armed) %>% 
    filter(is.na(armed)) %>% 
    .$n -> n_na

n_na / nrow(deaths)  # percentage of missing data

```

---

### human details

```{r age, echo=FALSE}
# age
deaths %>% 
    filter(!is.na(age)) %>% 
    ggplot(aes(age)) +
        geom_histogram(breaks=seq(0,110,10),fill=prim_color,color="black",alpha=0.8) +
        labs(title="age distribution",x="",y="") +
        scale_y_continuous(labels=scales::comma)
```

The age range represented in the data set is from **`r min(deaths$age,na.rm=TRUE)`** to **`r max(deaths$age,na.rm=TRUE)`**. The average age is **`r signif(mean(deaths$age,na.rm=TRUE),digits=2)`**.


```{r gender, echo=FALSE}
# gender
genders <- deaths %>%
    count(gender)

males <- genders$n[which(genders$gender=="Male")]
females <- genders$n[which(genders$gender=="Female")]
gender_nas <- genders$n[is.na(genders$gender)]

perc_male <- males/(males+females)
perc_na <- gender_nas/sum(genders$n)

genders %>% 
    mutate(gender=replace_na(gender,"Unknown")) %>% 
    ggplot(aes(gender,n)) +
        geom_col(fill=prim_color,color="black",alpha=0.8,width=0.8) +
        labs(title="gender",x="",y="") +
        scale_y_continuous(labels=scales::comma)

```

The people represented in the data were **`r signif(perc_male*100,digits=3)`%** males. Gender was unspecified for `r signif(perc_na*100,digits=2)`% of the people represented.


```{r capita_demographics, echo=FALSE}
us_pop_transpose <- us_pop %>% 
    pivot_longer(cols=Total:Hispanic,names_to="race",values_to="population")

deaths %>% 
    filter(!is.na(race),race!="Other") %>% 
    group_by(race) %>% 
    summarise(count=n(),.groups="drop") %>% 
    left_join(us_pop_transpose,by="race") %>% 
    mutate(rate=count/population) %>% 
    ggplot(aes(fct_reorder(race,desc(rate)),rate*1e6)) +
        geom_col(fill=prim_color,color="black",alpha=0.8) +
        labs(title="per capita police-involved deaths",subtitle="(all ages and all causes)",x="",y="(deaths per million)")
```


```{r capita_demographics_under20, echo=FALSE}
deaths %>% 
    filter(!is.na(race),race!="Other",age<=20) %>% 
    group_by(race) %>% 
    summarise(count=n(),.groups="drop") %>% 
    left_join(us_pop_transpose,by="race") %>% 
    mutate(rate=count/population) %>% 
    ggplot(aes(fct_reorder(race,desc(rate)),rate*1e6)) +
        geom_col(fill=prim_color,color="black",alpha=0.8) +
        labs(title="per capita police-involved deaths",subtitle="(under the age of 20)",x="",y="(deaths per million)")
```


```{r capita_demographics_notshot, echo=FALSE}
deaths %>% 
    filter(!is.na(race),race!="Other",manner_of_death!="Shot") %>% 
    group_by(race) %>% 
    summarise(count=n(),.groups="drop") %>% 
    left_join(us_pop_transpose,by="race") %>% 
    mutate(rate=count/population) %>% 
    ggplot(aes(fct_reorder(race,desc(rate)),rate*1e6)) +
        geom_col(fill=prim_color,color="black",alpha=0.8) +
        labs(title="per capita police-involved deaths",subtitle="(manner of death other than gunshot)",x="",y="(deaths per million)")
```


```{r age_demographics, echo=FALSE, fig.width=9}
deaths %>% 
    filter(!is.na(age),!is.na(race),race!="Other") %>% 
    group_by(race) %>% 
    summarise(average_age=mean(age),.groups="drop") %>% 
    ggplot(aes(fct_reorder(race,average_age),average_age)) +
        geom_col(color="black",fill=prim_color,alpha=0.8,width=0.5) +
        labs(title="age demographics",x="",y="average age") +
        geom_text(aes(label=signif(average_age,digits=3)),size=3.5,nudge_y=-2) +
        coord_flip()

```

---

### u.s. national trend (2000-2016)

```{r echo=FALSE, fig.width=10, fig.height=7}
deaths %>% 
    ggplot(aes(date)) +
        geom_histogram(aes(fill=manner_of_death),breaks=seq(as.Date("2000-01-01"),as.Date("2017-01-01"),"years"),
                       color="black",alpha=0.8) +
        labs(title="",x="",y="police-involved deaths",fill="manner of death") +
        annotate(geom="text",x=as.Date("2014-03-01"),y=2160,hjust="right",
                 label="increases in agressive policing?*",size=3.5
                 ) +
        annotate(geom="segment",
                 x=as.Date("2014-04-01"),xend=as.Date("2014-11-01"),
                 y=2150,yend=2120,
                 size=0.3
                 ) +
        scale_y_continuous(labels=scales::comma)

```

*Reference: Are Police Pulling the Trigger Too Fast? (All Things Considered)[^1]

[^1]:https://www.npr.org/2016/07/15/486150716/are-police-being-taught-to-pull-the-trigger-too-fast