---
title: "u.s. gun deaths (2014-2017)"
output: html_document
---
    

```{r setup, include=FALSE}
library(dslabs)
library(tidyverse)
library(lubridate)
library(stringr)
library(RColorBrewer)
library(ggsci)
library(ggrepel)
theme_set(theme_minimal())
knitr::opts_chunk$set(echo=FALSE,fig.width=9,fig.height=6)
```


```{r, include=FALSE}
# load murders dataset
data(murders)
murders <- as_tibble(murders)

# load dataset of guns incidents
gun_incidents <- as_tibble( read_csv("gun-violence-data_01-2013_03-2018.csv") )

# load state and region data file
states <- read.csv("../library/states.csv")

# clean, tidy and munge
gs <- gun_incidents %>% 
    select(id=incident_id,
           date,
           state,
           city=city_or_county,
           dead=n_killed,
           injured=n_injured,
           type=gun_type,
           nguns=n_guns_involved,
           age=participant_age,
           gender=participant_gender,
           name=participant_name,
           relates=participant_relationship
           ) %>% 
    mutate(state=as.factor(state),
           city=as.factor(city),
           gender=as.factor(gender),
           year=year(date),
           month=month(date),
           day=day(date)
           ) %>% 
    filter(between(year,2014,2017))  # 2013, 2018 data do not appear to include full year
            
```

    
The data represented is from the [Gun Violence Archive](https://www.gunviolencearchive.org/) provided by [James Ko](https://www.kaggle.com/jameslko/gun-violence-data) representing the **`r format(nrow(gs),big.mark=",")`** gun incidents recorded from 2014 to 2017.


```{r}
gs %>% 
    mutate(year=as.factor(year),
           year=fct_rev(year)
           ) %>% 
    group_by(state,year) %>% 
    summarise(dead_state=sum(dead),.groups="drop") %>%
    left_join(murders,by="state") %>%  # left join with murders data set to associate with population data
    select(-region,-total) %>% 
    left_join(states,by=c("state"="State")) %>% 
    rename(region=Region) %>% 
    mutate(region=as.factor(region)) %>% 
    select(-Division,-State.Code) %>%
    group_by(region,year) %>%
    summarise(dead_per_capita=mean(dead_state/population*1e5),.groups="drop") %>%
    ggplot(aes(x=reorder(region,dead_per_capita),y=dead_per_capita,fill=year)) +
        geom_bar(stat="identity",width=0.5,alpha=0.8,color="black") +
        coord_flip() +
        ylab("persons killed by gun violence (per 100,000)") + xlab("") +
        ggtitle("u.s. gun deaths per capita by region") +
        scale_fill_jco()

# bar graph of deaths per capita by region
```

Note:  2013 population data was used for convenience, so the actual numbers on the per capita chart vary somewhat from the actual, but the regional relationships are not expected to vary significantly from what is shown.

<br>

```{r}
# per capita gun ownership by region
gowners <- read.delim("raw-thoughtco-guns-by-state.txt")

gowners <- gowners %>% 
    mutate(state=as.character(State),
           guns_pcap=X..of.guns.per.capita,
           guns_reg=as.numeric(gsub(",","",as.character(X..of.guns.registered))),
           ) %>%
    select(state,guns_pcap,guns_reg) %>% 
    mutate(state=replace(state,state=="Washington D.C.","District of Columbia")) %>% 
    left_join(murders,by="state") %>%  # join with murders data frame for population
    mutate(state=as.factor(state)) %>% 
    select(-region,-total) %>%
    left_join(states,by=c("state"="State")) %>%  # join with states frame for region
    rename(region=Region) %>%
    select(-State.Code,-Division)

gowners %>% 
    group_by(region) %>% 
    summarise(guns_per_capita=sum(guns_reg)/sum(population),.groups="drop") %>%
    ggplot(aes(x=reorder(region,guns_per_capita),y=guns_per_capita*1e3)) +
        geom_bar(stat="identity",width=0.5,alpha=0.8,fill="#78c1f9",color="black") +
        coord_flip() +
        ylab("guns (per thousand)") + xlab("") +
        ggtitle("registered guns per capita by region")

```

<br>

```{r}
gs %>% 
    group_by(state) %>% 
    summarise(deaths=sum(dead),.groups="drop") %>%
    left_join(gowners,by="state") %>%
    mutate(deaths_capita=deaths/population) %>%
    filter(region %in% c("South","Northeast")) %>% 
    ggplot(aes(x=guns_reg/population*1e3,y=deaths_capita*1e5)) +
        geom_smooth(method="lm",formula=y~x,fill="#eeeeee",color="#cccccc") +
        geom_point(aes(fill=region),size=5,shape=21,stroke=1,alpha=0.8) +
        geom_text_repel(aes(label=state),
                   color="grey24",
                   point.padding=0.5) +
        scale_x_continuous(trans="log2") +
        scale_y_continuous(trans="log2") +
        scale_fill_jco() +
        ggtitle("gun deaths vs number of guns") +
        xlab("number of guns (per thousand)") +
        ylab("deaths (per 100,000)")

```

<br>

```{r}
# is the number of people who are killed or injured at once significant?
gs %>% 
    filter((dead!=0)|(injured!=0)) %>%
    mutate(mass=ifelse(dead>=4,TRUE,FALSE),
           dead=ifelse(dead==0,NA,dead),
           injured=ifelse(injured==0,NA,injured)
           ) %>% 
    select(state,dead,injured,mass) %>% 
    pivot_longer(cols=c("dead","injured"),
                 names_to="type",
                 values_to="count"
                 ) %>% 
    ggplot(aes(x=type,y=count)) +
        geom_hline(yintercept=3,linetype="dashed",color="grey") +
        geom_jitter(na.rm=TRUE,shape=21,fill="#78c1f9",stroke=1,size=2) +
        ggtitle("u.s. gun incidents") +
        xlab("") + ylab("")

```

Note:  In the United States, the term *mass killing* is defined as "3 or more killings in a single incident" (https://www.govinfo.gov/content/pkg/PLAW-112publ265/pdf/PLAW-112publ265.pdf).

<br>

```{r}
gs %>% 
    group_by(month) %>% 
    summarise(deaths=sum(dead),.groups="drop") %>%
    ggplot(aes(x=reorder(month.abb[month],month),y=deaths)) +
        geom_point(shape=21,fill="#fdc086",color="black",alpha=0.8,size=4.5,stroke=1) +
        xlab("") +
        ggtitle("u.s. gun deaths by month") +
        scale_y_continuous(labels=scales::comma)
    
```


```{r eval=FALSE}
state_gun_incidents <- gs %>% 
    group_by(state) %>% 
    summarise(count=n(),.groups="drop") %>% 
    left_join(murders,by="state") %>%  # left join with murders data set to associate with population data
    select(-region,-total) %>% 
    left_join(states,by=c("state"="State"))
    
state_gun_incidents %>% 
    ggplot(aes(population,count)) +
        geom_smooth(method="lm") +
        geom_point(aes(color=Region),size=4) +
        scale_x_continuous(trans="log10") +
        scale_y_continuous(trans="log10")
        
state_inc_model <- lm(state_gun_incidents$count ~ state_gun_incidents$population)

state_residuals <- resid(state_inc_model)

state_gun_incidents %>% 
    mutate(residuals=state_residuals) %>% 
    filter(Region %in% c("South","West")) %>% 
    ggplot(aes(population,residuals)) +
        geom_point(aes(fill=Region),size=5,shape=21,color="black",stroke=1) +
        scale_x_continuous(trans="log10",labels=scales::comma) +
        scale_y_continuous(labels=scales::comma) +
        geom_text_repel(aes(label=state),
                   color="grey24",
                   point.padding=0.5)

```

