---
title: "Objectives"
author: "Duane Edmonds"
date: '20220515'
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

# Objectives

This is the implementation of Objectives and Key Results ( OKRs ). The basic implementation is of a single objective datatype that could be represented in code as an custom object data structure cleverly using object-oriented program methodology to optimize the efficiency and beauty of the design and the code. That may be the direction I take as a poet.

On the other hand, there is another more relational approach that could become something that is perhaps just as delicious in a tidy and elegantly creative relational database implementation. Perhaps there's time for both. We'll start here. I'll choose MariaDB as the relational database management system. ( SQLite would be light and fast, but I want a more common standard-ish SQL interface. A mobile version would probably be done using SQLite. ) 

```{r objectives}
objectives <- 
  data.frame(
    id = TODO,
    description = TODO, 
    due_date = TODO, 
    comments = TODO, 
    rank = TODO, 
    sup_id = TODO
  )
```
